Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models

### 2012-03-05 revision/paper.tex
getting close to addressing all comments

% resolution issues with large data - reference - aloso long time series.

@@

@@

The glyph-map uses a small glyph, or icon, to represent multiple variables, or time values, at each location. Glyph maps derive from glyph plots, which were initially developed to display multivariate data. One glyph is drawn for each observation in the data set. Types of glyph plots include the star plots \citep{mayr:1877, kleiner:1981}, the semi-graphic displays \citep{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}. An glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.

@@

\item[GISTEMP] surface temperature data provided on $2^{\circ}$ x
  $2^{\circ}$ grid over the entire globe, measured monthly
  \citep{GISTEMP}. Ground station data was de-seasonalized,
  differenced from from the 1951-1980 temperature averages, and
  spatially averaged to obtain gridded temperature anomalies. For the
  purposes of this paper, we extracted the locations corresponding to
  the continental USA.

@@

A possible draw-back of the glyph-map, particularly if linear trend is displayed as an icon, is it may suffer from the Z\"ollner Illusion \citep{Zollner}, visible in Figure \ref{fig:gistemp-pred}, which makes straight lines look crooked.

@@

Each glyph is small, so there is not enough space for a full set of axes. Instead, minimal reference lines and boxes are incorporated. These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the data. The reference grid is a box, framing each icon, which represents the spatial grid. The reference line is a horizontal line at mid-range. Both help to read differences in slope and intercept. Slope is read by comparing the position of left and right endpoints on the frame, or by reading the angle between the data line and the reference line. Intercept is read from ``average'' position of the line in the box: is it near the top, or near the bottom? Figure~\ref{fig:ref-basic} demonstrates both guides. Reference elements are drawn in white, so they are minimally perceptible relative to the black data.

@@

For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle. Figures~\ref{fig:templates} and \ref{fig:gistemp-raw} show more examples.

@@

Alternatively, local scale can be used. In local scaling, the values within each location are scaled to range $[0, 1]$ before plot construction. This makes it easier to compare shape (ignoring amplitude), and draws attention to locations with large relative variation. Figure~\ref{fig:scaling} compares global scaling and local scaling for the smoothed temperature data from the data-expo data. On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location. Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, dramatic linear increases across the Andes, and decreases in the Gulf of Mexico.

>>>>

% resolution issues with large data - reference - also long time series - large spatial domain.

% order of the algorithm calculations

@@

% refer to the wrapping of times series in eden

@@

The glyph-map uses a small glyph, or icon, to represent multiple variables, or time values, at each location. Glyph maps derive from glyph plots, which were initially developed to display multivariate data. One glyph is drawn for each observation in the data set. Types of glyph plots include the star plots \citep{mayr:1877, FD01, kleiner:1981}, the semi-graphic displays \citep{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}. An glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.

@@

\item[GISTEMP] surface temperature data provided on $2^{\circ}$ x
  $2^{\circ}$ grid over the entire globe, measured monthly
  \citep{GISTEMP} from 1880-2011. Ground station data was de-seasonalized,
  differenced from from the 1951-1980 temperature averages, and
  spatially averaged to obtain gridded temperature anomalies. For the
  purposes of this paper, we extracted the locations corresponding to
  the continental USA.

@@

A possible draw-back of the glyph-map, particularly if linear trend is displayed as an icon, is it may suffer from the Z\"ollner Illusion \citep{Zollner}, which makes straight lines look crooked (Section \ref{sec:large-data} has an example).

@@

Each glyph is small, so there is not enough space for a full set of axes. Instead, minimal reference lines and boxes are incorporated. These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the data. The reference grid is a box, framing each icon, which represents the spatial grid. The reference line is a horizontal line at mid-range. Both help to read differences in slope and intercept. Slope is read by comparing the position of left and right endpoints on the frame, or by reading the angle between the data line and the reference line. Intercept is read from ``average'' position of the line in the box: is it near the top, or near the bottom? Figure~\ref{fig:ref-basic} demonstrates both guides. Reference elements are drawn in white, so they are minimally perceptible relative to the black data. Average temperatures were calculated for each month, and these were passed to the glyph calculations to produce the glyph-maps.

@@

For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle (see Figure~\ref{fig:templates}).

@@

Alternatively, local scale can be used. In local scaling, the values within each location are scaled to range $[0, 1]$ before plot construction. This makes it easier to compare shape (ignoring amplitude), and draws attention to locations with large relative variation. Figure~\ref{fig:scaling} compares global scaling and local scaling for the smoothed temperature data from the data-expo data. (Generalized additive model using smooth terms as described in \citet{wood:2006} were used.)  On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location. Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, linear increases across the Andes, and decreases in the Gulf of Mexico.
